---
title: "R Notebook"
author: "Karen Mazidi"
output:
  pdf_document: default
  html_notebook: default
---

# Data exploration in R with the Titanic data set.

### Load the data

Next we use the read.csv() function to read a csv in a subdirectory called data. Once  you read in the data you will see that it has 1310 observations of 14 variables. We run the str() structure function to get a peek at the data.

```{r}
df <- read.csv("data/titanic3.csv", na.strings="NA", stringsAsFactors=FALSE, header=TRUE)
str(df)
```
### Data cleaning

The read.csv() function is a bit aggressive about making things factors. Generally if the column contains character data, it tries to make it a factor. Sometimes this makes sense, sometimes it does not. 

We can change a column to a factor with as.factor() or change a column to integer with as.integer() as shown next.

```{r}
df$pclass <- as.factor(df$pclass)
df$sex <- factor(df$sex, levels=c("male", "female"))
```

### Factors

Factors are stored internally as integer vectors but also have a character representation for human readability. We can use contrasts() to find out more about a factor column.

The contrasts for pclass shows that we need 2 variables to encode 3 classes. The base case will be class 1. R will create 2 dummy variables for classes 2 and 3. We will see the importance of these when we get to machine learning.


```{r}
contrasts(df$pclass)
contrasts(df$sex)
```

That's all for now. We will revisit the Titanic data later when we explore classification algorithms: learning how to predict who survived and who didn't based on demographic data in the file.
