---
title: "Text Processing with tm"
author: "Karen Mazidi"
output:
  pdf_document: default
  html_notebook: default
---

Amazon review data from Kaggle: https://www.kaggle.com/bittlingmayer/amazonreviews

### Load data and package tm

```{r}
library(tm)
reviews <- read.csv("data/reviews.csv", header=TRUE, stringsAsFactors=F)
```

### counts for ratings

Rating = 1 means 1 or 2 stars; Rate=2 mean 4 or 5 stars; 3 stars were ignored

About 46% were low and 54% high. 

```{r}
low_ratings <- nrow(reviews[reviews$Rating == 1,])
high_ratings <- nrow(reviews[reviews$Rating == 2,])
```


### Make a simple corpus

```{r}
am_corpus <- Corpus(VectorSource(reviews$Review))
inspect(am_corpus[1])
```

### Preprocess

```{r}
am_clean <- tm_map(am_corpus, content_transformer(tolower))
am_clean <- tm_map(am_clean, removeNumbers)
am_clean <- tm_map(am_clean, removePunctuation)
am_clean <- tm_map(am_clean, removeWords, stopwords())
am_clean <- tm_map(am_clean, stripWhitespace)
```

### Make Document Term Matrix

```{r}
am_dtm <- DocumentTermMatrix(am_clean)
am_dtm
```

### Divide into test and train

```{r}
set.seed(1234)
i <- sample(nrow(reviews), 0.75*nrow(reviews), replace=FALSE)
# labels
train_labels <- reviews[i, 1]
test_labels <- reviews[-i, 1]
# data
train <- am_clean[i]
test <- am_clean[-i]


```

### Ignore rare words

```{r}
freq_words <- findFreqTerms(am_dtm, 5)
train <- DocumentTermMatrix(train, control=list(dictionary=freq_words))
test <- DocumentTermMatrix(test, control=list(dictionary=freq_words))
inspect(train[50:55, 200:209])
```

### Create binary matrix

```{r}
convert_count <- function(x) {
  y <- ifelse(x>0, 1, 0)
  y <- factor(y)
  y
}

train <- apply(train, 2, convert_count)
test <- apply(test, 2, convert_count)
```

### Naive Bayes

```{r}
library(e1071)
nb1 <- naiveBayes(train, factor(train_labels))
```

### Evaluate on test data


We only got 54% accuracy. Further improvements might be made by removing proper nouns and stemming, but we leave that for future work.

```{r}
pred <- predict(nb1, newdata=test)
table(pred, test_labels)
mean(pred == test_labels)

```

