---
title: "Text Processing with RTextTools"
author: "Karen Mazidi"
output:
  pdf_document: default
  html_notebook: default
---

The RTextTools package integrates text processing and machine learning. Read more about it in [this paper](https://journal.r-project.org/archive/2013-1/collingwood-jurka-boydstun-etal.pdf).

We are going to look at using RTextTools for processing the amazon reviews data.

### Load the data

```{r}
library(RTextTools)
reviews <- read.csv("data/reviews.csv", header=TRUE, stringsAsFactors=F)
```

### Create a document term matrix

This uses the tm package under the hood.

```{r}
dtm <- create_matrix(reviews$Review, language="english", removeNumbers=TRUE,
                     removeStopwords=TRUE, stemWords=TRUE, removeSparseTerms=.998)
```

### Create a container

The container will hold train and test observations as well as labels.

```{r}
container <- create_container(dtm, reviews$Rating, trainSize=1:3000,
                              testSize=3001:4139, virgin=FALSE)
```

### Train model

There are several algorithms to choose from, we just selected 3 of them.

```{r}
svm <- train_model(container, "SVM")
glmnet <- train_model(container, "GLMNET")
maxent <- train_model(container, "MAXENT")
```

### Classify

Now apply the models to the test data.

```{r}
svm_classify <- classify_model(container, svm)
glmnet_classify <- classify_model(container, glmnet)
maxent_classify <- classify_model(container, maxent)
```

### Analytics

Interpreting the results. 

```{r}
analytics <- create_analytics(container, cbind(
      svm_classify, glmnet_classify, maxent_classify))
summary(analytics)
```
### Create ensemble aggreement

Calculate coverage, the percentage of cases on which the n cases agree, for n >= 1, 2, 3 models. 

```{r}
create_ensembleSummary(analytics@document_summary)
```
