---
title: "Introduction to dplyr and ggplot2"
author: "Karen Mazidi"
output:
  pdf_document: default
  html_notebook: default
editor_options: 
  chunk_output_type: console
---
## Explore dplyr

### load dplyr and data

The dplyr package was designed to be efficient with large data but we will demonstrate the basic features with a smaller data set from package mlbench.

```{r, message=FALSE}
library(dplyr)
library(mlbench)
data("PimaIndiansDiabetes2")
```

### tbl

A tbl "tibble" is a data frame with enhanced features. Now when we type the name at the console we get a neater display of our data, one page at a time. 

```{r}
df <- tbl_df(PimaIndiansDiabetes2)
rm(PimaIndiansDiabetes2)
df
```

### glimpse

The glimpse function is similar to str but can handle bigger data more efficiently.

```{r}
glimpse(df)
```

### 5 functions of dplyr

* select - remove columns
* mutate - create new columns from the data
* filter - remove rows
* arrange - rearrange rows
* summarize or summarise - summary statistics

These functions do not change the original data but return a new object. The functions assume that the data is already  "tidy" -- observations in rows, features in columns. 


### select -- used to select columns

Select a couple of columns to print. Notice it doesn't run off the screen, you get a screen's worth at a time. 

```{r}
print(select(df, diabetes, pregnant))
```

### mutate -- used to add columns

Add a column that is a binary factor indicating if glucose is above average for the population.

```{r}
mutate(df, glucose_high = as.factor(ifelse(glucose>mean(glucose, na.rm=TRUE), 1, 0)))
```

### filter -- used to remove rows

We replace df with a df that filtered out rows with NAs in the glucose or mass columns.

```{r}
df <- filter(df, !is.na(glucose), !is.na(mass))
glimpse(df)
```

### arrange -- arrange rows based on content

Arrange rows based on bmi as stored in the mass column. 

```{r}
arrange(df, mass)  # ascending order
arrange(df, desc(mass)) # descending order
```

### summarize - a more powerful summary

Get summary statistics on mass.

```{r}
summarize(df, min=min(mass), max=max(mass), sd(mass))
summarize(df, n_diabetic = sum(diabetes=="pos"), n_not_diabetic = sum(diabetes=="neg"))
```


### pipes - work similar to unix pipes

Pipes make code easier to read and let you make several commands in one neat group of lines instead of nesting functions in the typical R fashion. The pipe operator %>% comes from package magrittr but dplyr automatically loads it. 

```{r}
df %>%
  group_by(diabetes) %>%
  summarize(n_diabetic = n())

```

## Explore ggplot2

Hadley Wickham developed ggplot2 in 2005, inspired by a grammar of graphics developed by Leland Wildinson in 1999. The ggplot2 functions are much more powerful than standard R graphs but also slower. 

We have a short example below showing important components of building a ggplot. First we specify the data, then the aesthetics which are how the data is represented, followed by the geometry and finally labels.

```{r, message=FALSE}
library(ggplot2)
# load data
library(mlbench)
data("PimaIndiansDiabetes2")
df <- PimaIndiansDiabetes2[complete.cases(PimaIndiansDiabetes2[]),]

ggplot(df, aes(x=mass, y=glucose)) +
  geom_point() +
  labs(title="Glucose and BMI", x="BMI", y="Glucose")
```
Next we add some color and a smoothing line which helps us see a trend in the data. By default the smoothing line has a shadow around it which specifies the 95

```{r}
ggplot(df, aes(x=mass, y=glucose)) +
  geom_point(pch=20, color='blue', size=1.5) +
  geom_smooth(method='lm', color='red', linetype=2) +
  labs(title="Glucose and BMI", x="BMI", y="Glucose")
```
### facet_grid

```{r}
df$glucose_high <- factor(ifelse(df$glucose>mean(df$glucose, na.rm=TRUE), 1, 0),
                           levels=c(0,1), labels=c("Glucose High","Glucose Not High"))
df$insulin_high <- factor(ifelse(df$insulin>mean(df$insulin, na.rm=TRUE), 1, 0),
                           levels=c(0,1), labels=c("Insulin High","Insulin Not High"))

ggplot(df, 
  aes(x=df$mass, y=df$age, shape=diabetes, col=pregnant)) +
  geom_point(size=2) +
  facet_grid(df$glucose_high~df$insulin_high)

```
### histogram

```{r}
ggplot(df, aes(x=mass)) +
  geom_histogram(fill="cornsilk4")
```

### boxplot and rug

```{r}
ggplot(df, aes(x=diabetes, y=mass)) +
  geom_boxplot(notch=TRUE) +
  geom_point(position="jitter", color="cornflowerblue", alpha=.5) +
  geom_rug(color="cornflowerblue")
```

### density plot

```{r}
ggplot(df, aes(x=mass, fill=diabetes)) +
  geom_density(alpha=0.5)
```
### bubble chart

```{r}
ggplot(df, 
       aes(x=mass, y=glucose, size=pregnant)) +
  geom_point(shape=21, fill="cornflowerblue")
```

### grid

```{r}
library(gridExtra)
p1 <- ggplot(df, aes(x=insulin_high)) + geom_bar(fill="cornflowerblue") 
p2 <- ggplot(df, aes(x=glucose_high)) + geom_bar(fill="cornflowerblue")
grid.arrange(p1, p2, ncol=2)
```

