---
title: "Neural Network Classification"
author: "Karen Mazidi"
output:
  pdf_document: default
  html_notebook: default
editor_options: 
  chunk_output_type: inline
---

Classification on the PimaIndiansDiabetes2 data set in package mlbench.

### Load data, divide into train and test

```{r}
library(mlbench)
data("PimaIndiansDiabetes2")
df <- PimaIndiansDiabetes2[complete.cases(PimaIndiansDiabetes2[]),]
set.seed(1234)
i <- sample(1:nrow(df), 0.75*nrow(df), replace=FALSE)
train <- df[i,]
test <- df[-i,]
```

### Logistic regression

Performing logistic regression as a baseline. Accuracy on the test set is .765.


```{r}
glm1 <- glm(diabetes~., data=train, family=binomial)
probs <- predict(glm1, newdata=test, type="response")
pred1 <- ifelse(probs>0.5, "pos", "neg")
mean(pred1==test$diabetes)
```
### Preprocessing

We need to scale the predictors. 

```{r}
scaled <- scale(df[,-9])
df_scaled <- data.frame(cbind(scaled, df$diabetes))
colnames(df_scaled) <- colnames(df)
df_scaled$diabetes <- factor(df_scaled$diabetes)
train_scaled <- df_scaled[i,]
test_scaled <- df_scaled[-i,]
```

### More preprocessing

The neuralnet() function does not handle factors. We have to replace the target factor with a one-hot encoding of diabetes. 

```{r}
class_column <- train_scaled$diabetes
n <- length(class_column)
x <- matrix(0, n, length(levels(class_column)))
x[(1L:n) + n * (unclass(class_column) - 1L)] <- 1
dimnames(x) <- list(names(class_column), levels(class_column))
train_scaled <- cbind(train_scaled[,1:8], x)    
names(train_scaled) <- c(names(train_scaled)[1:8], "neg", "pos")
```


### Create the formula

```{r}
n <- names(train)
f <- as.formula(paste("neg + pos ~", paste(n[!n %in% "diabetes"], collapse = " + ")))
```

### Build the neural network

```{r}
library(neuralnet)
set.seed(1234)
nnet1 <- neuralnet(f, data=train_scaled, hidden=c(5,2), threshold=0.1, act.fct = "logistic", linear.output=FALSE, lifesign = "minimal")
plot(nnet1)
```


### Evaluate on test set

The mean accuracy was 0.776, one percent higher than logistic regression.

```{r}
nnet.results <- compute(nnet1, test_scaled[,1:8])
pred <- max.col(nnet.results$net.result)
mean(test_scaled$diabetes==pred)
```




