---
title: "Beijing PM2.5 Data"
output:
  pdf_document: default
  html_notebook: default
  html_document:
    df_print: paged
---

Hourly data of PM2.5 from the US Embassy in Beijing. PM2.5 is a measure of particulate matter that have a diameter of less than 2.5 micrometers. They are an important measure of air quality for humans. 

The data was downloaded from the UCI Machine Learning Repository and is calleed the [Beijing PM2.5 Data Set](https://archive.ics.uci.edu/ml/datasets/Beijing+PM2.5+Data#)

### Load data

Loading the data and restricting to complete cases leaves about 41K observations. We will remove the No, day, and hour columns. Remaining columns are the year and month, pm2.5, temperature, pressure, combined wind direction. cumulated wind speed, cumulated hours of snow and cumulated hours of rain.

```{r}
library(keras)
df <- read.csv("PRSA_data.csv", header=TRUE)
df <- df[complete.cases(df), c(3, 6:13)]
head(df)
str(df)

N <- nrow(df)
p <- ncol(df)
t <- 2

X <- df[, -t]
Y <- df[, t]
```


### Train/test split

```{r}
set.seed(1234)
i <- sample(1:nrow(df), 0.8*nrow(df), replace=FALSE)
X_train <- data.matrix(X[i,])
Y_train <- Y[i]
X_test <- data.matrix(X[-i,])
Y_test <- Y[-i]
```


### normalize data

```{r}
means <- apply(X_train, 2, mean)
stdvs <- apply(X_train, 2, sd)
X_train <- scale(X_train, center=means, scale=stdvs)
X_test <- scale(X_test, center=means, scale=stdvs)
```



### Try neural network

```{r}
# build a model
model <- keras_model_sequential()
model %>%
  layer_dense(units=16, activation='relu', input_shape = dim(X_train)[[2]]) %>%
  layer_dense(units=16, activation='relu') %>%
  layer_dense(units=1)

model %>% compile(
  loss = 'mse',
  optimizer = 'rmsprop',
  metrics = c("mae")
)

model %>% fit(X_train, Y_train, epochs=100, batch_size=100, verbose=0)

results <- model %>% evaluate(X_test, Y_test, verbose=0)
results$mean_absolute_error 

# how do you get predictions?
pred <- predict(model, X_test)
cor(pred, Y_test)  # 0.7
mse <- mean((pred - Y_test)^2)   # 4211
sqrt(mse)  # 64.9
mae <- mean(abs(pred - Y_test))  # 44.4

```




