---
title: "04-2-ggplot2"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    theme: united
---

### Code Accompanying ***The Machine Learning Handbooks***, Volume I, Chapter 4

#### Book pdf is available on the GitHub repo: https://github.com/kjmazidi/Machine_Learning_3rd_edition

###### Copyright KJG Mazidi

### Introduction

This notebook demonstrates plotting with ggplot2.

Packages imported in this notebook:

- ggplot2 (grammar of graphics)
- mlbench for the PimaIndiansDiabetes2 data set
- gridExtra for arranging graphs

```{r, warning = FALSE, message = FALSE}
if (!require(tidyverse)){
  install.packages("tidyverse")
}
library("tidyverse")

if (!require(mlbench)){
  insyes
  tall.packages("mlbench")
}
library("mlbench")

if (!require(gridExtra)){
  insyes
  tall.packages("gridExtra")
}
library("gridExtra")
```

### gg is for grammar of graphics

There are 7 grammatical elements in ggplot2, the first 3 of these are essential to getting something plotted:

* data - the data being plotted should be the first argument, or specify data=...
* aesthetics - the scales onto which we plot; use aes() to specify at least x= and y= if needed as well as other parameters for customization
* geometries - visual elements such as points, lines, etc.
* facets - for plotting multiples
* statistics - representations to aid understanding
* coordinates - space on which data will be plotted
* themes - you can customize your own theme to use over and over

### load tidyverse and some data

Loading the diabetes data set from package mlbench.

```{r}
data("PimaIndiansDiabetes2")

tb <- as_tibble(PimaIndiansDiabetes2)
```


## Explore ggplot2

Hadley Wickham developed ggplot2 in 2005, inspired by a grammar of graphics developed by Leland Wildinson in 1999. The ggplot2 functions are much more powerful than standard R graphs but also slower. 

We have a short example below showing important components of building a ggplot. First we specify the data, then the aesthetics which are how the data is represented, followed by the geometry and finally labels.

You will notice that these graphs produce many warnings about data observations. In this notebook we did no data cleaning before the graphical data exploration which is why those warnings pop up. 

```{r}
ggplot(tb, aes(x=mass, y=glucose)) +
  geom_point() +
  labs(title="Glucose and BMI", x="BMI", y="Glucose")
```

Next we add some color and a smoothing line which helps us see a trend in the data. By default the smoothing line to highlight the trend in the data

```{r}
ggplot(tb, aes(x=mass, y=glucose)) +
  geom_point(pch=20, color='blue', size=1.5) +
  geom_smooth(method='lm', color='red', linetype=2) +
  labs(title="Glucose and BMI", x="BMI", y="Glucose")
```

### informative graph


```{r}
ggplot(tb, 
  aes(x=tb$mass, y=tb$age, shape=diabetes, col=pregnant)) +
  geom_point(size=2) +
  labs(x="BMI",    y="Age")
```



### facet_grid

* filter out rows with NAs in glucose or insulin
* create 2 new factor columns, glucose_high and insulin_high
* plot

The facet grid for 2 binary variables has 4 windows for all combinations. 

```{r}
tb <- filter(tb, !is.na(glucose), !is.na(insulin))

tb <- mutate(tb, glucose_high = factor(ifelse(tb$glucose>mean(tb$glucose), 1, 0)))
tb <- mutate(tb, insulin_high = factor(ifelse(tb$insulin>mean(tb$insulin), 1, 0)))

ggplot(tb, 
  aes(x=mass, y=age, shape=diabetes, col=pregnant)) +
  geom_point(size=2) +
 facet_grid(glucose_high~insulin_high)       
```


### histogram

```{r}
ggplot(tb, aes(x=mass)) +
  theme_bw() +
  geom_histogram(fill="cornsilk4")
```

### boxplot and rug

```{r}
ggplot(tb, aes(x=diabetes, y=mass)) +
  geom_boxplot(notch=TRUE) +
  geom_point(position="jitter", color="cornflowerblue", alpha=.5) +
  geom_rug(color="cornflowerblue")
```

### density plot

```{r}
ggplot(tb, aes(x=mass, fill=diabetes)) +
  geom_density(alpha=0.5)
```
### bubble chart

```{r}
ggplot(tb, 
       aes(x=mass, y=glucose, size=pregnant)) +
  geom_point(shape=21, fill="cornflowerblue")
```

### grid

```{r}
library(gridExtra)
p1 <- ggplot(tb, aes(x=insulin_high)) + geom_bar(fill="cornflowerblue") 
p2 <- ggplot(tb, aes(x=glucose_high)) + geom_bar(fill="cornflowerblue")
grid.arrange(p1, p2, ncol=2)
```

