---
title: "Tidyverse Demo"
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
---

## Code Accompanying **The Machine Learning Handbooks**, Volume I, Chapter 4

This code demonstrates the tidyverse packages and functions.

Packages needed in this notebook:

```{r}
if (!require(tidyverse)){
  install.packages("tidyverse")
}
library("tidyverse")

if (!require(mlbench)){
  insyes
  tall.packages("mlbench")
}
library("mlbench")
```



### Create a tibble

```{r}
# use a mlbench data frame
data("PimaIndiansDiabetes2")

tb <- as_tibble(PimaIndiansDiabetes2)
tb

# remove the data frame to free up memory
rm(PimaIndiansDiabetes2)
```

A glimpse is a view similar to str.

```{r}
glimpse(tb)
```

### The dplyr package

Some dply functions work on columns. These are demonstrated below.

#### select()

Select a subset of columns. The select() function returns a tibble but it was not saved and will be discarded after the glimpse is output.

```{r}
select(tb, diabetes, pregnant) %>%
  glimpse
```

or:

```{r}
tb %>%
  select(diabetes, pregnant) %>%
  glimpse
```

#### mutate()

The mutate() function can create new columns from old ones.

```{r}
tb <- tb %>%
       mutate(glucose_high = factor(
         ifelse(glucose>mean(glucose, na.rm=TRUE), 1, 0)))

tb[1:5, c(2, 10)]
```

We can also use mutate to delete a column by setting it to NULL.

```{r}
tb <- tb %>%
  mutate(glucose_high = NULL)

names(tb)
```

#### rename()

Rename a column.

```{r}
tb <- rename(tb, blood_pressure = pressure)
```

#### filter()

The filter function can select rows.

```{r}
tb <- filter(tb, !is.na(glucose), !is.na(mass))
glimpse(tb)
```
#### arrange()

The following code arranges the rows by mass in descending order.

```{r}
arrange(tb, desc(mass))
```

### summarize

The summarize function computes statistical summaries of the data.

```{r}
tb %>%
  summarize(min=min(mass), max=max(mass), sd=sd(mass))
```

Another example:

```{r}
tb %>%
  summarize(num_diabetic = sum(diabetes=="pos"), num_healthy = sum(diabetes=="neg"))
```
### count 

A simpler way to get the same information as above:

```{r}
tb %>% count(diabetes, sort = TRUE)
```



```{r}

```


### group_by

```{r}
tb %>%
  group_by(diabetes) %>%
  summarize(median_BMI = median(mass, na.rm=TRUE))
```

